---
title: "Inequalities in violent death among men, by state, race, and firearms use"
output:
  word_document:
    reference_docx: reference.docx
bibliography: biblio.bib
---

Corinne A Riddell, PhD

Sam Harper, PhD

Magdalena Cerdá, PhD

Jay S Kaufman, PhD

CAR, SH, and JK: Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada.

MC: Violence Prevention Research Program, Department of Emergency Medicine, University of California, Davis.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>. Phone: +1-438-863-0402

##### PAGE BREAK

#Key Points

(75 to 100 words or less)

**Question:** How much do the inequalities in homicide and suicide rates between black and white men vary by US state and firearms use?

**Finding:** There was substantial variation in the difference in rates of firearm-related homicide and suicide between blacks and whites. Black men experienced 
between 5 and 51 additional homicides per 100,000, while white men experienced between -2 and 16 additional suicides per 100,000 men, across US states.

**Meaning:** Policies aimed at reducing inequalities in violence, and violence overall, should be targeted towards states exhibiting inequalities in these deaths.

##### PAGE BREAK
#Abstract
350 word limit

**Importance:** It is unknown whether and how much the difference in excess deaths due to homicide among black men and suicide among white men varies by US state. Such knowledge could inform social policy aimed towards reducing social disparities in health.

**Objective:** (i) Estimate the absolute inequality between non-Hispanic black and white men in firearm- and non-firearm-related homicide and suicide across US states; (ii) Examine the relationships between state gun ownership prevalence, firearm-related homicide, and firearm-related suicide, separately for black and white men.

**Design:** Cross-sectional study.

**Setting:** 50 states and the District of Columbia, 2008 to 2014.

**Participants:** Non-Hispanic black and non-Hispanic white males, all ages.

**Exposure:** State gun prevalence (%)

**Main Outcomes and Measures:** Absolute risks and racial risk differences in firearm- and nonfirearm-related homicide and suicide for black and white men.

**Results:** 71,217 homicides and 191,909 suicides were included in this analysis. Wide variation in the difference in rates of firearm-related homicide and suicides between black and white men were identified. Across states, black men experienced between 5 and 51 additional homicides per 100,000 men per year, with Missouri and the Midwestern states of Michigan, Pennsylvania, Wisconsin, and Indiana having more than 40 additional homicides per year among black men. White men experienced between -2 and 16 additional suicides per 100,000 men, with the largest inequalities measured in Southern and Western states, and smallest inequalities in densely populated Northeastern states. Descriptive analyses suggest that shared explanatory factors including state gun prevalence explain the variation in firearm-related homicide and suicide among white men, but not among black men, who exhibit excessive variation in homicide rates.

**Conclusions and Relevance:** Guns feature differently in the lives of black and white men. The large variation in the risks of violence and their racial inequalities across states puts into the spotlight states where policies are most urgently needed to reduce rates of violence and their racial inequalities. Future research linking these variations with the social, cultural, and policy environments are needed to identify settings that support reduced rates of violence for all Americans.

##### PAGE BREAK

```{r, echo=F, warning=F, message=F}
library(tidyverse)
library(ggrepel)
library(viridis)
```

```{r gun_ownership_data, echo=F, warning=F, message=F}
gun.house.prev <- c(52.2, 59.8, 32.3, 58.8, 20.1, 34.6, 18.1, 26.3, 25.2,
                    40.3, 10.2, 55.7, 20.7, 38.5, 45.7, 42.8, 47.7, 45.0,
                    40.3, 21.7, 11.5, 40.8, 41.2, 54.6, 44.2, 62.6, 45.4,
                    34.0, 31.0, 11.4, 39.7, 18.5, 39.4, 56.2, 34.0, 46.5,
                    39.8, 35.1, 12.4, 43.3, 59.9, 46.6, 37.1, 44.8, 43.8,
                    37.5, 34.0, 58.5, 43.0, 65.5)

gun.data <- data.frame("ST" = state.abb, "gun.house.prev" = gun.house.prev)

#ggplot(gun.data, aes(gun.house.prev)) + geom_histogram(bins = 20, col = "white") 
#ggplot(gun.data, aes(gun.house.prev)) + geom_density()
#quantile(gun.data$gun.house.prev, c(0.33, 0.66))

gun.data <- gun.data %>% arrange(gun.house.prev) %>% 
  mutate(gun.house.prev.category = cut(gun.house.prev, 
                                       breaks = c(0, 20, 35, 44.9, 66), 
                                       labels = c("10.2%-19.9%", "20.0%-34.9%", "35.0%-44.9%", "45.0%-65.5%")))
```

```{r load_data_age_std_males, echo=F, warning=F, message=F}
#homicide
NHWM_homicide_FA <- read.table("./Data/agestd/Homicide_Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHWM_homicide_NonFA <- read.table("./Data/agestd/Homicide_Non-Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_homicide_FA <- read.table("./Data/agestd/Homicide_Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_homicide_NonFA <- read.table("./Data/agestd/Homicide_Non-Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

#suicide
NHWM_suicide_FA <- read.table("./Data/agestd/Suicide_Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHWM_suicide_NonFA <- read.table("./Data/agestd/Suicide_Non-Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_suicide_FA <- read.table("./Data/agestd/Suicide_Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_suicide_NonFA <- read.table("./Data/agestd/Suicide_Non-Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")
```

```{r load_data_age_std_females, echo=F, warning=F, message=F}
#homicide
NHWF_homicide_FA <- read.table("./Data/agestd/Homicide_Firearms_White_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHWF_homicide_NonFA <- read.table("./Data/agestd/Homicide_Non-Firearms_White_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBF_homicide_FA <- read.table("./Data/agestd/Homicide_Firearms_Black_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBF_homicide_NonFA <- read.table("./Data/agestd/Homicide_Non-Firearms_Black_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

#suicide
NHWF_suicide_FA <- read.table("./Data/agestd/Suicide_Firearms_White_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHWF_suicide_NonFA <- read.table("./Data/agestd/Suicide_Non-Firearms_White_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBF_suicide_FA <- read.table("./Data/agestd/Suicide_Firearms_Black_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBF_suicide_NonFA <- read.table("./Data/agestd/Suicide_Non-Firearms_Black_NH_Females_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")
```

```{r tidy_fx, echo=F, warning=F, message=F}
tidy_fx <- function(data, state, death, population, crude, adjusted, race, intent, mechanism) {
  dat2 <- data[, c(state, death, population, crude, adjusted)]
  names(dat2)[2] <- paste0(death, ".", intent, ".", race)
  names(dat2)[3] <- paste0(population, ".", race)
  names(dat2)[4] <- paste0("crude.", intent, ".", race)
  names(dat2)[5] <- paste0("adjusted.", intent, ".", race)
  dat2$type = mechanism
  
  return(dat2)
}
```

```{r tidy_male_data, echo=F}
NHWM_homicide_FA <- tidy_fx(data = NHWM_homicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "homicide", "Firearm")
NHWM_homicide_NonFA <- tidy_fx(data = NHWM_homicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "homicide", "Non-Firearm")
NHWM_suicide_FA <- tidy_fx(data = NHWM_suicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "suicide", "Firearm")
NHWM_suicide_NonFA <- tidy_fx(data = NHWM_suicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "suicide", "Non-Firearm")

NHBM_homicide_FA <- tidy_fx(data = NHBM_homicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "homicide", "Firearm")
NHBM_homicide_NonFA <- tidy_fx(data = NHBM_homicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "homicide", "Non-Firearm")
NHBM_suicide_FA <- tidy_fx(data = NHBM_suicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "suicide", "Firearm")
NHBM_suicide_NonFA <- tidy_fx(data = NHBM_suicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "suicide", "Non-Firearm")

NHBM_homicide<- rbind(NHBM_homicide_FA, NHBM_homicide_NonFA)
NHBM_suicide <- rbind(NHBM_suicide_FA, NHBM_suicide_NonFA)
NHWM_homicide <- rbind(NHWM_homicide_FA, NHWM_homicide_NonFA)
NHWM_suicide <- rbind(NHWM_suicide_FA, NHWM_suicide_NonFA)
rm(NHBM_homicide_FA, NHBM_homicide_NonFA, NHBM_suicide_FA, NHBM_suicide_NonFA,
   NHWM_homicide_FA, NHWM_homicide_NonFA, NHWM_suicide_FA, NHWM_suicide_NonFA)
```

```{r tidy_female_data, echo = F}
NHWF_homicide_FA <- tidy_fx(data = NHWF_homicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "homicide", "Firearm")
NHWF_homicide_NonFA <- tidy_fx(data = NHWF_homicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "homicide", "Non-Firearm")
NHWF_suicide_FA <- tidy_fx(data = NHWF_suicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "suicide", "Firearm")
NHWF_suicide_NonFA <- tidy_fx(data = NHWF_suicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "suicide", "Non-Firearm")

NHBF_homicide_FA <- tidy_fx(data = NHBF_homicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "homicide", "Firearm")
NHBF_homicide_NonFA <- tidy_fx(data = NHBF_homicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "homicide", "Non-Firearm")
NHBF_suicide_FA <- tidy_fx(data = NHBF_suicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "suicide", "Firearm")
NHBF_suicide_NonFA <- tidy_fx(data = NHBF_suicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "suicide", "Non-Firearm")

NHBF_homicide<- rbind(NHBF_homicide_FA, NHBF_homicide_NonFA)
NHBF_suicide <- rbind(NHBF_suicide_FA, NHBF_suicide_NonFA)
NHWF_homicide <- rbind(NHWF_homicide_FA, NHWF_homicide_NonFA)
NHWF_suicide <- rbind(NHWF_suicide_FA, NHWF_suicide_NonFA)
rm(NHBF_homicide_FA, NHBF_homicide_NonFA, NHBF_suicide_FA, NHBF_suicide_NonFA,
   NHWF_homicide_FA, NHWF_homicide_NonFA, NHWF_suicide_FA, NHWF_suicide_NonFA)
```

```{r merge_data_males, echo=F}
#merge all the data
Males_AllAges_0814 <- merge(NHBM_homicide, NHBM_suicide %>% select(-Population.black), by = c("ST", "type")) %>%
  mutate(crude.RD.hom.su.black = crude.homicide.black - crude.suicide.black,
         adj.RD.hom.su.black = adjusted.homicide.black - adjusted.suicide.black)

Males_AllAges_0814 <- merge(Males_AllAges_0814, NHWM_homicide, by = c("ST", "type")) %>%
  mutate(crude.RD.homicide = crude.homicide.black - crude.homicide.white,
         adj.RD.homicide = adjusted.homicide.black - adjusted.homicide.white)

Males_AllAges_0814 <- merge(Males_AllAges_0814, NHWM_suicide %>% select(-Population.white), by = c("ST", "type")) %>%
  mutate(crude.RD.suicide = crude.suicide.black - crude.suicide.white,
         adj.RD.suicide = adjusted.suicide.black - adjusted.suicide.white)

Males_AllAges_0814 <- Males_AllAges_0814 %>% 
  mutate(crude.RD.hom.su.white = crude.homicide.white - crude.suicide.white, 
         adj.RD.hom.su.white = adjusted.homicide.white - adjusted.suicide.white)
```

```{r merge_data_females, echo=F}
Females_AllAges_0814 <- merge(NHBF_homicide, NHBF_suicide %>% select(-Population.black), by = c("ST", "type")) %>%
  mutate(crude.RD.hom.su.black = crude.homicide.black - crude.suicide.black,
         adj.RD.hom.su.black = adjusted.homicide.black - adjusted.suicide.black)

Females_AllAges_0814 <- merge(Females_AllAges_0814, NHWF_homicide, by = c("ST", "type")) %>%
  mutate(crude.RD.homicide = crude.homicide.black - crude.homicide.white,
         adj.RD.homicide = adjusted.homicide.black - adjusted.homicide.white)

Females_AllAges_0814 <- merge(Females_AllAges_0814, NHWF_suicide %>% select(-Population.white), by = c("ST", "type")) %>%
  mutate(crude.RD.suicide = crude.suicide.black - crude.suicide.white,
         adj.RD.suicide = adjusted.suicide.black - adjusted.suicide.white)

Females_AllAges_0814 <- Females_AllAges_0814 %>% 
  mutate(crude.RD.hom.su.white = crude.homicide.white - crude.suicide.white, 
         adj.RD.hom.su.white = adjusted.homicide.white - adjusted.suicide.white)
```

```{r add_factor_sorting_males, echo = F}
#used for plotting -- to arrange the states on the y-axis according to the magnitude of the black-white inequality
levels.firearm.homicide <- Males_AllAges_0814 %>% filter(type == "Firearm") %>%
  transmute(ST.order.RD.homicide = reorder(ST, adj.RD.homicide))

# Males_AllAges_0814 %>% 
#   filter(is.na(adjusted.homicide.black) == T & type == "Firearm") %>% 
#   select(ST, adjusted.homicide.white)

Males_AllAges_0814$ST.order.RD.homicide <- factor(Males_AllAges_0814$ST, 
                                                  levels = c("HI", "SD", "NH", "ND", "VT", "ID", "ME", "WY", "MT",
                                                             levels(levels.firearm.homicide$ST.order.RD.homicide)[1:42]))

levels.firearm.suicide <- Males_AllAges_0814 %>% filter(type == "Firearm") %>%
  transmute(ST.order.RD.suicide = reorder(ST, -adj.RD.suicide))

# Males_AllAges_0814 %>% 
#   filter(is.na(adjusted.suicide.black) == T & type == "Firearm") %>% 
#   select(ST, adjusted.suicide.white)

Males_AllAges_0814$ST.order.RD.suicide <- factor(Males_AllAges_0814$ST, 
                                                  levels = c("RI", "HI", "IA", "NH", "ME", "SD", "ND", "VT", "UT", "ID", "MT", "WY",
                                                             levels(levels.firearm.suicide$ST.order.RD.suicide)[1:39]))

```

```{r add_factor_sorting_females, echo = F}
#used for plotting -- to arrange the states on the y-axis according to the magnitude of the black-white inequality
levels.firearm.homicide <- Females_AllAges_0814 %>% filter(type == "Firearm") %>%
  transmute(ST.order.RD.homicide = reorder(ST, adj.RD.homicide))

# Females_AllAges_0814 %>% 
#   filter(is.na(adjusted.homicide.black) == T & type == "Firearm") %>% 
#   select(ST, adjusted.homicide.white)

Females_AllAges_0814$ST.order.RD.homicide <- factor(Females_AllAges_0814$ST, 
                                                  levels = levels(levels.firearm.homicide$ST.order.RD.homicide))

# Males_AllAges_0814$ST.order.RD.homicide <- factor(Males_AllAges_0814$ST, 
#                                                   levels = c("HI", "SD", "NH", "ND", "VT", "ID", "ME", "WY", "MT",
#                                                              levels(levels.firearm.homicide$ST.order.RD.homicide)[1:42]))

levels.firearm.suicide <- Females_AllAges_0814 %>% filter(type == "Firearm") %>%
  transmute(ST.order.RD.suicide = reorder(ST, -adj.RD.suicide))

# Females_AllAges_0814 %>% 
#   filter(is.na(adjusted.suicide.black) == T & type == "Firearm") %>% 
#   select(ST, adjusted.suicide.white)

# Females_AllAges_0814$ST.order.RD.suicide <- factor(Females_AllAges_0814$ST, 
#                                                   levels = c("RI", "HI", "IA", "NH", "ME", "SD", "ND", "VT", "UT", "ID", "MT", "WY",
#                                                              levels(levels.firearm.suicide$ST.order.RD.suicide)[1:39]))

Females_AllAges_0814$ST.order.RD.suicide <- factor(Females_AllAges_0814$ST, 
                                                  levels = levels(levels.firearm.suicide$ST.order.RD.suicide))

```

```{r add_variables_males, echo=F}
num.years = 2014-2008+1

Males_AllAges_0814 <- Males_AllAges_0814 %>% 
  mutate(average.pop.white = Population.white/num.years,
         average.pop.black = Population.black/num.years)

Males_AllAges_0814$gun.level <- "Middle"
Males_AllAges_0814$gun.level[Males_AllAges_0814$ST %in% c("LA", "UT", "OK", "IA", "TN","KY", "AL", 
                                                          "MS", "ID","ND", "WV", "AR", "AK", "SD", "MT", "WY")] <- "High"
Males_AllAges_0814$gun.level[Males_AllAges_0814$ST %in% c("HI", "NJ", "MA", "RI", "CT", "NY")] <- "Low"

Males_AllAges_0814$gun.level <- factor(Males_AllAges_0814$gun.level, levels = c("Low", "Middle", "High"))

#dim(Males_AllAges_0814)
Males_AllAges_0814 <- merge(Males_AllAges_0814, gun.data, by = "ST", all.x = T)
#dim(Males_AllAges_0814)
```

```{r add_variables_females, echo=F}
Females_AllAges_0814 <- Females_AllAges_0814 %>% 
  mutate(average.pop.white = Population.white/num.years,
         average.pop.black = Population.black/num.years)

Females_AllAges_0814$gun.level <- "Middle"
Females_AllAges_0814$gun.level[Females_AllAges_0814$ST %in% c("LA", "UT", "OK", "IA", "TN","KY", "AL", 
                                                          "MS", "ID","ND", "WV", "AR", "AK", "SD", "MT", "WY")] <- "High"
Females_AllAges_0814$gun.level[Females_AllAges_0814$ST %in% c("HI", "NJ", "MA", "RI", "CT", "NY")] <- "Low"

Females_AllAges_0814$gun.level <- factor(Females_AllAges_0814$gun.level, levels = c("Low", "Middle", "High"))

#dim(Females_AllAges_0814)
Females_AllAges_0814 <- merge(Females_AllAges_0814, gun.data, by = "ST", all.x = T)
#dim(Females_AllAges_0814)
```

#Introduction

```{r WISQARS-data-intro, echo=F}
#source: WISQARS fatal injury mapping tool
#note:I uses homicide, not homicide and legal intervention -- this changes the numbers slightly, not materially. 

#HOMICIDE relative risks
nh.black.homicide.rate.agestd <- 33.58 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.homicide.rate.crude <- 35.53 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.homicide.rate.agestd <- 3.46 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.homicide.rate.crude <- 3.45 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.homicide.black.white.agestd <- nh.black.homicide.rate.agestd/nh.white.homicide.rate.agestd
rr.homicide.black.white.crude <- nh.black.homicide.rate.crude/nh.white.homicide.rate.crude

#SUICIDE relative risks
nh.black.suicide.rate.agestd <- 9.64 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.suicide.rate.crude <- 9.32 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.suicide.rate.agestd <- 24.49 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.suicide.rate.crude <- 26.01 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.suicide.white.black.agestd <- nh.white.suicide.rate.agestd/nh.black.suicide.rate.agestd
rr.suicide.white.black.crude <- nh.white.suicide.rate.crude/nh.black.suicide.rate.crude

```

```{r WISQARS-data-intro-2, echo=F}

risk.homicide.nh.male <- 8.33 #per 100k, non-hispanic blacks + whites, 2008-2014, males
risk.homicide.nh.male.firearm <- 6.17 #per 100k, non-hispanic blacks + whites, 2008-2014, males
prop.homicide.nh.male.firearm <- (risk.homicide.nh.male.firearm/risk.homicide.nh.male)*100

risk.suicide.nh.male <- 22.42 #per 100k, non-hispanic blacks + whites, 2008-2014, males
risk.suicide.nh.male.firearm <- 12.87 #per 100k, non-hispanic blacks + whites, 2008-2014, males
prop.suicide.nh.male.firearm <- (risk.suicide.nh.male.firearm/risk.suicide.nh.male)*100
```

```{r WISQARS-data-intro-3, echo=F}
#source: WISQARS fatal injury mapping tool
#note:I uses homicide, not homicide and legal intervention -- this changes the numbers slightly, not materially. 

#FIREARM HOMICIDE relative risks
nh.black.homicide.rate.agestd.fa <- 27.58 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.homicide.rate.crude.fa <- 29.57 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.homicide.rate.agestd.fa <- 2.04 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.homicide.rate.crude.fa <- 2.01 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.homicide.black.white.agestd.fa <- nh.black.homicide.rate.agestd.fa/nh.white.homicide.rate.agestd.fa
rr.homicide.black.white.crude.fa <- nh.black.homicide.rate.crude.fa/nh.white.homicide.rate.crude.fa

#FIREARM SUICIDE relative risks
nh.black.suicide.rate.agestd.fa <- 5.33 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.suicide.rate.crude.fa <- 5.04 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.suicide.rate.agestd.fa <- 14.06 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.suicide.rate.crude.fa <- 15.27 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.suicide.white.black.agestd.fa <- nh.white.suicide.rate.agestd.fa/nh.black.suicide.rate.agestd.fa
rr.suicide.white.black.crude.fa <- nh.white.suicide.rate.crude.fa/nh.black.suicide.rate.crude.fa

```

Non-Hispanic black males are `r round(rr.homicide.black.white.agestd,1)` times more likely to die from homicide than non-Hispanic white males in the United States, while white males are `r round(rr.suicide.white.black.agestd,1)` times more likely to die from suicide [@wisqars_data]. `r round(prop.homicide.nh.male.firearm)`% of these homicides and `r round(prop.suicide.nh.male.firearm)`% of these suicides involved firearms, and the relative inequalities in violent deaths are especially marked for firearm-related homicide and slightly increased for firearm-related suicide; Black males are `r round(rr.homicide.black.white.agestd.fa,1)` times more likely to die of a firearm-related homicide than white males, while white males are `r round(rr.suicide.white.black.agestd,1)` times as likely to die from firearm-related suicide [@wisqars_data]. 

While these racial inequalities in violent deaths are well-established for the United States overall, it is unknown whether and how much they vary by state. Large variation between racial inequalities in spatially contiguous or otherwise similar states may reflect differences in the social, cultural, and policy environments worthy of further investigation. Looking forward, this information is the first step towards implementing targeted policies aimed at reducing violence and inequalities where they are most urgently needed. Reduction of firearm-related mortality overall is an important public health pursuit in its own right, however, reductions in inequalities are also important, especially if they are considered inequities associated with more adverse social conditions linked to race or ethnicity [@arcaya2015inequalities]. 

Across the US, states with the highest levels of household gun ownership have higher rates of firearm-related suicide and homicide [@webster_book_table]. Ecologic studies comparing US states have estimated positive associations between rates of gun ownership and firearm-related homicide [@miller2002rates, @miller2007state, @siegel2013relationship] and suicide rates,[@miller2013firearms, @miller2007household] while case-control studies how found positive associations between having a gun in the home and an individual’s risk of homicide occurring at home,[@kellermann1993gun] and with the overall risk of suicide [@kellermann1992suicide]. If the dynamics of where homicides or suicides occur (at home vs. outside of the home) vary by race, then the relationship between gun prevalence and firearm fatalities may be differential by race. This information is currently unknown but is important to inform whether policies aimed at reducing gun prevalence may have differential effects among blacks and whites.

[[To put into a comment box: This is a little weak, I think. There is one paper by Kubrin and Wadsworth [@kubrin2009explaining] that looks at predictors of suicide by race (separately) before and after adjustment for gun availability. They find that disadvantage is biggest predictor of suicide for both blacks and whites but the effect is attenuated for blacks when firearm availability is added to the model but not for whites. They motivate this well by talking about motivation for suicide and opportunity for suicide as two pieces of the causal pie. Gun availability is a proxy for gun possession and therefore a proxy for opportunity for suicide. Motivation is the other piece and varies by race (perhaps). So maybe the argument should be: If motivation for violence (homicide or suicide) varies by race, then the relationship between gun prevalence and fatalities may also differ.???]]

In this study, we aimed to: (1) Estimate the absolute inequality between non-Hispanic black and white men in firearm- and non-firearm-related homicide and suicide across US states; (2) Examine the relationships between state gun ownership prevalence, firearm-related homicide, and firearm-related suicide, separately for black and white men.

#Methods

We used the Center for Disease Control and Prevention’s Web-based Injury Statistics Query and Reporting System (WISQARS) Fatal Injury Mapping tool to abstract the age-adjusted number of firearm fatalities for males in every state during 2008 to 2014 for non-Hispanic whites and non-Hispanic blacks, by intent (homicide or suicide), and mechanism (firearms or non-firearms). Throughout this paper we refer to this population as men since the vast majority of deaths occurred after childhood. The data was suppressed by WISQARS for race-state strata with 10 or fewer actual deaths between 2008 and 2010, and could not be used in these analyses. We computed the average annual fatality rate according to race, intent, and firearms usage by dividing the number of fatalities by the total population estimate (i.e., the summation of population estimates over 2008 to 2014) and we calculated the absolute inequality (risk difference) in estimates between black and white men For homicide, we subtract the fatality rate of white men from that of black men, and for suicide these rates are reversed, such that the estimated risk differences are positive and interpreted as the number of additional deaths to black (white) men due to homicide (suicide) associated with the heightened fatality rate experienced by black (white) men. We did not calculate confidence intervals because there was no sampling, and thus no sampling error. To reflect varying population sizes among states and by race, we size the data markers according to race-specific average annualized population size. 

To examine the relationships between gun prevalence, firearm-related homicide, and firearm-related suicide by race, we used the last available state-level estimates of household firearm ownership from the National Behaviourial Risk Factor Surveillance System (BRFSS) from 2004. The BRFSS is an annual telephone survey on the non-institutionalized adult population that asks respondents about their health-related risk behaviors, chronic health conditions, and use of preventive services. In 2001, 2002, and 2004, the BRFSS asked, “Are any firearms kept in or around your home?” 293,882 people responded to this question and estimates of household gun ownership prevalence (%) by state were computed previously and ranged between 10.2% in Hawaii and 65.5% in Wyoming [@monuteaux2015firearm]. Given that individuals may lie about gun ownership, and that state gun prevalence may have shifted since 2004, we grouped states into four categories of gun prevalence: (i) 10% - 19.9%, (ii) 20% - 34.9%, (iii) 35% - 44.9%, and (iv) 45% - 66%, to provide a fair proxy for current levels of ownership. 

We then visually assessed the relationships between state gun prevalence, firearm-related homicide, and firearm-related suicide by graphing the rate of firearm-related homicide as a function of firearm-related suicide and shading the points by category of gun prevalence, separately for black men and white men. Correlation between rates of firearm-related homicide and suicide would imply that state conditions that generate variation in homicide also generate variation in suicide, such as cultural, social, or economic conditions, including state gun prevalence or the firearm policy environment. Thus, the purpose of this descriptive exercise is to illustrate whether states with relatively high rates of firearm-related homicide also have relatively high rates of firearm-related suicide, and if this pattern exists within race strata and is associated with gun prevalence. 

The analysis was performed in R version 3.3.2,[@r_language] with all code available on Github (link: https://github.com/corinne-riddell/InjuryDisparity).

#Results

```{r results_statistics, echo=F}
stats <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adj.RD.homicide, na.rm = T)), max = round(max(adj.RD.homicide, na.rm = T)))

stats2 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.homicide.white, na.rm = T)), max = round(max(adjusted.homicide.white, na.rm = T)))

white.pop.dc <- Males_AllAges_0814 %>% filter(ST == "DC" & type == "Firearm") %>% select(Population.white)

stats3 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.homicide.black, na.rm = T)), max = round(max(adjusted.homicide.black, na.rm = T)))

tib1 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(sum.deaths.homicide.black = sum(Deaths.homicide.black, na.rm = T), 
                                                    sum.deaths.homicide.white = sum(Deaths.homicide.white, na.rm = T), 
                                                    sum.deaths.suicide.black = sum(Deaths.suicide.black, na.rm = T), 
                                                    sum.deaths.suicide.white = sum(Deaths.suicide.white, na.rm = T))


```

There were `r sum(tib1$sum.deaths.homicide.black) + sum(tib1$sum.deaths.homicide.white)` homicides and `r sum(tib1$sum.deaths.suicide.black) + sum(tib1$sum.deaths.suicide.white)` suicides during 2008 to 2014 included in this analysis. Homicide deaths were supressed for black men in 11 states (Alaska (non-firearm only), New Mexico (non-firearm only) Montana, Wyoming, Maine, Idaho, Vermont, North Dakota, New Hampshire, South Dakota, and Hawaii), and for white men in DC. Suicide deaths were supressed in 12 states for black men (Montana, Wyoming, Maine, Idaho, Vermont, North Dakota, New Hampshire, South Dakota, Hawaii (firearm only), Utah (firearm only), Iowa (firearm only), and Rhode Island (firearm only)).

**Figure 1** displays the heightened rates of homicide among black men compared to white men, according to state and firearms use. The black-white difference in homicide fatality rates for men varied strikingly by state, between `r stats$min[1]` and `r stats$max[1]` per 100,000 per year, and were driven by marked variation in firearm homicides to black men across states. States with the smallest black firearm homicide rates saw fewer than 15 deaths per 100,000 per year over the time period, compared to more than 40 deaths per 100,000 per year in states with the highest rates. Homicide rates for white men showed relatively little variation, between `r stats2$min[1]` and `r stats2$max[1]` deaths per 100,000 per year. The states with the highest rates of firearm homicide for black men (e.g., Missouri, Michigan, Illinois, Indiana) also had the largest differences in rates between blacks and whites, signaling settings that were most dangerous for blacks but relatively safer for whites. The District of Columbia also had a high rate of firearm homicide for blacks, but its rate for white men was supressed. Given its abstracted cumulative population denominator of `r white.pop.dc`, this implies a rate $\leq$ 10 per `r white.pop.dc`, or `r round((10/white.pop.dc)*100000, 1)` per 100,000, implying that DC was also relatively dangerous for black men, but not for white men.

Georgia, New York, Florida, Texas, and California are the states with the largest black populations. Among them, Texas exhibited the smallest black-white difference in firearm-related homicide (13 additional deaths to black men per 100,000), which was also substantially lower than the differences estimated for neighboring states like Louisiana (+37/100,000), Oklahoma (+29/100,000), and Arkansas (+29/100,000). Georgia (+16/100,000) and Florida (+22/100,000) had differences close to nearby South and North Carolina (+19 and +17 per 100,000, respectively), while having lower rates than Alabama (+27/100,000) and Tennessee (+28/100,000). New York's difference (+18/100,000) was similar to that estimated for Massachusettes (+17/100,000) and Connecticut (+22/100,000), but was substantially lower than the rates of neighboring New Jersey (+30/100,000) and Pennsylvania (+41/100,000). California exhibited the largest black-white difference (+30/100,000) in firearm-related homicide among the five states with the largest black populations, which was larger than the rates estimates for contiguous states like Nevada (+17/100,000) and Oregon (+10/100,000), although these states had relatively small black populations.

Non-firearm homicides occurred relatively rarely, and maintained the general pattern of heightened rates among black men compared to white men. While black men experienced annual rates that varied between `r stats3$min[2]` and `r stats3$max[2]` per 100,000 by state, white rates varied between `r stats2$min[2]` and `r stats2$max[2]` per 100,000. Across the US, firearm homicide rates were higher than non-firearm rates for black men, and were more than 5.3 times as high in all states in the East North Central (Wisconsin, Illinois, Indiana, Ohio, and Michigan), and in Missouri, Louisiana, Delaware, and Nebraska.

```{r results_statistics_suicide, echo=F}
stats4 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adj.RD.suicide, na.rm = T)),
                                                    max = round(max(adj.RD.suicide, na.rm = T)))

stats5 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.suicide.white, na.rm = T)),
                                                    max = round(max(adjusted.suicide.white, na.rm = T)))

stats6 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.suicide.black, na.rm = T)),
                                                    max = round(max(adjusted.suicide.black, na.rm = T)))

stats7 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.suicide.white, na.rm = T)),
                                                    max = round(max(adjusted.suicide.white, na.rm = T)))
```

In contrast to homicide, suicide fatality rates were higher among white men compared to black men, independent of firearms usage, with the exception of firearm-related suicide in DC (**Figure 2**). The white-black difference in suicides involving a firearm varied from `r -(stats4$max[1])` per 100,000 in the District of Columbia to `r -(stats4$min[1])` per 100,000 per year in Missouri, although differences could not be estimated for 12 states with 10 or fewer firearm-related suicides among blacks. The annual rate of suicide involving firearms varied more markedly for white men than for black men, between `r stats5$min[1]` per 100,000 in DC and `r stats5$max[1]` per 100,000 in Wyoming for white men, compared to a range of `r stats6$min[1]` per 100,000 in Massachusetts to `r stats6$max[1]` per 100,000 in New Mexico among black men. 

Seven of the ten states with the largest white-black differences (i.e., risk differences between 12 and 16 per 100,000) in firearm-related suicide are located in the South, including Mississippi, Arkansas, West Virginia, Alabama, Texas, South Carolina, and Louisiana, while the remaining three are in the West, including Nevada, Arizona, and Colorado. Along with DC, the mid-Atlantic states of Massachusetts, New Jersey, Connecticut, and New York had the smallest white-black differences (<4 per 100,000) in firearm-related suicide. Importantly, these states also tended to have the lowest firearm-related suicide rates for both black men and white men, implying that the states safest (in terms of suicide risk) for black men and white men are the same. 

The rates of black- and white- non-firearm-related suicides and the difference between them varied substantially less than those involving firearms. For white men, between `r stats7$min[2]` per 100,000 in Alaska and `r stats7$max[2]` per 100,000 in Hawaii non-firearm-related suicides were committed each year on average across the states, compared to `r stats6$min[2]` per 100,000 in Alabama and `r stats6$max[2]` per 100,000 in Utah among black men.

The five states with jointly low rates of firearm-related homicide and suicide among white men (i.e., Massachusetts, New Jersey, Rhode Island, New York, and Connecticut) had the lowest category of gun prevalence (**Figure 3**). Furthermore, many of the states with the highest joint violence rates belonged to the highest category of gun prevalence, while most states located in the middle of the graph belonged to the middle categories of gun prevalence. Notably, several states that belonged to the highest gun prevalence category had relatively low rates of firearm-related homicide, including Iowa, Nebraska, the Dakotas, Idaho, Montana, and Wyoming. Generally speaking, the strong linear relationship between firearm-related homicide and suicide among white men suggests that the characteristics that generate variation in homicide and suicide were similar for white men.

Among black men, states with the lowest prevalence of guns had the lowest rates of firearm-related suicide, and below-average rates of firearm-related homicide (**Figure 4**). States with the highest joint rates of violence (i.e., Mississippi, Missouri, Oklahoma) had heightened rates of gun prevalence, although several Southern states with relatively high gun prevalence had relatively low rates of firearm-related homicide. The non-existent relationship between firearm-related homicide and suicide among black men implies that the set of characteristics that generated variation in firearm-related homicide among black men are different from the characteristics that generated variation in firearm-related suicide.

#Discussion

Using data from all 50 US states and DC, our study illustrates striking variation in the inequality in firearm-related homicide, largely driven by differences among states in homicide rates for black men. While Southern states have relatively high levels of state gun prevalence, these states saw relatively low rates of homicide for black men, in contrast to Midwestern states where the risk of homicide was twice as high.  Disentangling the factors that make the South relatively safe for black men and the Midwest relatively unsafe (while very safe for white men) is of high priority in order to reduce health inequalities and foster communities that offer equal safety independent of race. Sociologists have found that structural disadvantage (i.e., poverty, joblessness, family disruption),[@Wilson_book] racial residential segregation,[@Massey_book, @ousey2009exploring, @shihadeh1996segregation, @peterson1993racial] and participation in drug markets [@blumstein1995youth, @fryer2013measuring] are implicated in increased risk of homicide among black men relative to white men. Thus, future research incorporating these neighborhood characteristics could elucidate whether the wide variation in inequality among states is explained by these factors. 

White men are at increased risk for firearm-related suicide in every state where comparisons could be made (except for DC). Variation across states in suicide for white men was large, although to a lesser extent than the variation in homicide see for black men. State gun prevalence was strongly tied to firearm-related suicide for both black and white men, and is the most lethal means linked to suicide completion [@spicer2000suicide, @miller2004epidemiology]. While we did not look at alcohol in this study, a history of alcoholism is a risk factor for suicide [@moscicki1997identification, moscicki2001epidemiology], and varies by race. During 2005 -- 2006, alcohol was found in the systems of 25% of white men tested and 15% of black men, with 16% of the white men being described as alcohol dependent at the time of suicide compared to 7% of black men [https://www.cdc.gov/mmwr/preview/mmwrhtml/mm5823a1.htm]. Given that white men have higher rates of alcohol abuse than black men, alcohol use might be implicated both in the direction of the white-black inequality and its variation across states.  

This study is subject to several limitations. We considered only fatal outcomes, and missed attempted suicides and homicides. Given that suicide attempts are more than 80% fatal,[@miller2004epidemiology, @spicer2000suicide] and homicide attempts are approximately 20% fatal [@cook2017constant], studying only fatal outcomes misses the bulk of morbidity endured during nonfatal assaults and less of the morbidity endured in failed suicide attempts. We were limited to constrasting age-standardized mortality rates, but would have rather contrasted age-specific rates, had there been more statistical power. We did not examine women because their risks of homicide and suicide are much lower than for men, and data for black women were suppressed for more than half of the states. Finally, these results are purely descriptive and make no claims regarding causal mechanisms. Our purpose was to describe how much inequalities in violence vary across the United States in order to provide a foundation for future research linking these differences to variations in the social, cultural, and environmental settings that black and white communities are based, including variations in firearm policy and access to firearms.

Firearms feature differently in the lives of black and white men in the United States. The large variation in the homicide and suicide rates and their racial differences across states place a spotlight on states where policies are most urgently needed to reduce rates of violence, and their racial inequalities. This research provides the first step in measuring these inequalities across states, which can be further studied to identify the policies, and social and cultural factors that support reduced rates of violence for all Americans.





Since 1999, the rates of suicide increased for white men from 20.2 to 25.8 per 100,000, as it did among all other race-sex strata with the exception of black men who saw a small reduction from 10.5 to 9.7 per 100,000 [https://www.cdc.gov/nchs/data/hestat/suicide/rates_1999_2014.htm]. While suicide for white men had traditionally been highest among the elderly, the increasing trend in the age-standardized rate was largely due to an increase in suicide for men aged 45-64, from 20.8 in 1999 to 29.7 in 2014. Thus,  


>>white men suicide rate highest among the elderly >85 yo, but for black men it is among younger black men [shaffer 1994 ref for suicide rate worsening in black teens]. check if rate for black men young is still increasing. some say it has to do with increasing gun possession.
>> 1980-1990 the suicide rate for black teenagers rose -- hypothesized link to access to guns


and is also often found in the system of victims at the time of their deaths. 











 Firearm-related suicide inequalities varied across states but to a lesser degree, and was driven by variation in suicide rates for white men. States that were the most dangerous for black men in terms of their risk of homicide were very safe for white men, and included Missouri and the Midwestern states of Michigan, Wisconsin, Illinois, and Indiana. For firearm-related suicide, the states with the largest inequalities were from the South and the West. The states with the lowest inequalities in firearm-related suicide had relatively low rates for black and white men alike, including the Northeastern states of New York, Connecticut, New Jersey, Massachusetts, and the District of Columbia. 

stopped here: Given that such states also tend to have high gun availability, residents may live in settings that combine increased risk for suicide attempts with easy access to firearms, the most lethal means for suicide completion [@spicer2000suicide, @miller2004epidemiology]. Such settings can be compared to Northeastern states like New York, Massachusetts, Connecticut and New Jersey, which have the lowest rates of suicide for both black and white men and the fewest firearms. These states also have numerous restrictive gun laws,[@siegel2017firearm] pointing to a multi-faceted environment that has led to the lowest risks for suicide in the country.

For white men, there is a strong linear relationship between firearm-related homicide and suicide that suggests a shared set of factors affecting both outcomes. The positive association with state gun prevalence is also pronounced. This finding is aligned with the body of research linking gun prevalence or gun ownership to an individual's risk of suicide or homicide victimization (references from intro). Interestingly, there were several states with medium to high gun prevalences, average rates of firearm-related suicide, but reduced rates of firearm-related homicide, such as Minnesota, Wisconsin, Oregon, Colorado, and Utah. 

The relationship between firearm-related homicide, suicide, and gun prevalence is markedly different for black men. There is no discernible relationship between firearm-related homicide and suicide, suggesting that the factors leading to firearm-related homicide among black men differ from those leading to firearm-related suicide. This may be due to an additional set of factors exclusively affecting either homicide or suicide. 


- something about not wanting to narrow the gap by increasing risk of violence in either group. Specifically, we are interested in gaps because they might point to disparities in the lived experience by race.


##### PAGE BREAK

###Figure 1: Difference in age-standardized homicide fatalities rates for males by state, firearms, and race, 2008-2014

```{r figure1, fig.width=5, fig.height=6.5, echo = F, warning=F}
ggplot(Males_AllAges_0814, aes(y = ST.order.RD.homicide, x = adjusted.homicide.white)) + 
  geom_segment(aes(x = adjusted.homicide.white, xend = adjusted.homicide.black, 
                   y = ST.order.RD.homicide, yend = ST.order.RD.homicide), 
               lwd = 2, col = "navy", alpha = 0.5) +
  geom_point(aes(col = "NH White", size = average.pop.white)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1)
  geom_point(aes(x = adjusted.homicide.black, col = "NH Black", size = average.pop.black)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1),
  scale_color_manual(values = c("navy", "#fd8d3c")) +
  scale_alpha_continuous(guide = F) +
  scale_size_area(breaks = c(16600, 100000, 500000, 1500000, 3000000, 7000000)) +
  ylab("") +
  xlab("Homicide mortality per 100k") + 
  theme_bw() + 
  guides(color = guide_legend(title = " "), size = guide_legend(title = "Population")) + 
  theme(legend.position = c(0.8, 0.25), 
        plot.title = element_text(size = 10),
        legend.background = element_rect(fill=alpha('white', 0.2)),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12)) + 
   geom_text(data = Males_AllAges_0814 %>% filter(is.na(adj.RD.homicide) == F),
             aes(x = adjusted.homicide.black + 8, label = paste0("+", round(adj.RD.homicide))),
             size = 3, col = "navy") +
  facet_wrap( ~ type) +
  scale_x_continuous(limits = c(0, 70)) 
```

```{r figure1-females, fig.width=5, fig.height=6.5, echo = F, warning=F, eval = F}
ggplot(Females_AllAges_0814, aes(y = ST.order.RD.homicide, x = adjusted.homicide.white)) + 
  geom_segment(aes(x = adjusted.homicide.white, xend = adjusted.homicide.black, 
                   y = ST.order.RD.homicide, yend = ST.order.RD.homicide), 
               lwd = 2, col = "navy", alpha = 0.5) +
  geom_point(aes(col = "NH White", size = average.pop.white)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1)
  geom_point(aes(x = adjusted.homicide.black, col = "NH Black", size = average.pop.black)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1),
  scale_color_manual(values = c("navy", "#fd8d3c")) +
  scale_alpha_continuous(guide = F) +
  scale_size_area(breaks = c(16600, 100000, 500000, 1500000, 3000000, 7000000)) +
  ylab("") +
  xlab("Homicide mortality per 100k") + 
  theme_bw() + 
  guides(color = guide_legend(title = " "), size = guide_legend(title = "Population")) + 
  theme(legend.position = c(0.8, 0.25), 
        plot.title = element_text(size = 10),
        legend.background = element_rect(fill=alpha('white', 0.2)),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12)) + 
   geom_text(data = Females_AllAges_0814 %>% filter(is.na(adj.RD.homicide) == F),
             aes(x = adjusted.homicide.black + 1, label = paste0("+", round(adj.RD.homicide))),
             size = 3, col = "navy") +
  facet_wrap( ~ type) +
  scale_x_continuous(limits = c(0, 10)) 
```

##### PAGE BREAK

###Figure 2: Difference in suicide fatalities rates for males by state, firearms, and race, 2008-2014

```{r figure2, fig.width=5, fig.height=6.5, echo = F, warning=F}
ggplot(Males_AllAges_0814, aes(y = ST.order.RD.suicide, x = adjusted.suicide.white)) + 
  geom_segment(aes(x = adjusted.suicide.white, xend = adjusted.suicide.black, 
                   y = ST.order.RD.suicide, yend = ST.order.RD.suicide), 
               lwd = 2, col = "#fd8d3c", alpha = 0.5) +
  geom_point(aes(col = "NH White", size = average.pop.white)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1)
  geom_point(aes(x = adjusted.suicide.black, col = "NH Black", size = average.pop.black)) + #, alpha = ifelse(death.homicide.black < 20, 0.7, 1)
  scale_color_manual(values = c("navy", "#fd8d3c")) +
  scale_alpha_continuous(guide = F) +
  scale_size_area(breaks = c(16600, 100000, 500000, 1500000, 3000000, 7000000)) +
  ylab("") + xlab("Suicide mortality per 100k") + theme_bw() + 
  guides(color = guide_legend(title = " "), size = guide_legend(title = "Population")) + 
  theme(legend.position = c(0.90, 0.27), 
        plot.title = element_text(size = 10),
        legend.background = element_rect(fill=alpha('white', 0.2)),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12)) + 
  scale_x_continuous(limits = c(0, 35)) +
  #scale_x_continuous(limits = c(0, 60), labels = seq(0, 60, 20), breaks = seq(0, 60, 20)) +
  geom_text(data = subset(Males_AllAges_0814, is.na(adj.RD.suicide)==F), 
            aes(x = adjusted.suicide.white + 4, label = ifelse(ST != "DC", paste0("+", round(-adj.RD.suicide)), round(-adj.RD.suicide))), size = 3, col = "#fd8d3c") +
  facet_wrap(~type)
```

```{r figure2-females, fig.width=5, fig.height=6.5, echo = F, warning=F, eval = F}
ggplot(Females_AllAges_0814, aes(y = ST.order.RD.suicide, x = adjusted.suicide.white)) + 
  geom_segment(aes(x = adjusted.suicide.white, xend = adjusted.suicide.black, 
                   y = ST.order.RD.suicide, yend = ST.order.RD.suicide), 
               lwd = 2, col = "#fd8d3c", alpha = 0.5) +
  geom_point(aes(col = "NH White", size = average.pop.white)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1)
  geom_point(aes(x = adjusted.suicide.black, col = "NH Black", size = average.pop.black)) + #, alpha = ifelse(death.homicide.black < 20, 0.7, 1)
  scale_color_manual(values = c("navy", "#fd8d3c")) +
  scale_alpha_continuous(guide = F) +
  scale_size_area(breaks = c(16600, 100000, 500000, 1500000, 3000000, 7000000)) +
  ylab("") + xlab("Suicide mortality per 100k") + theme_bw() + 
  guides(color = guide_legend(title = " "), size = guide_legend(title = "Population")) + 
  theme( plot.title = element_text(size = 10),
        legend.background = element_rect(fill=alpha('white', 0.2)),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12)) + 
  scale_x_continuous(limits = c(0, 7)) +
  #scale_x_continuous(limits = c(0, 60), labels = seq(0, 60, 20), breaks = seq(0, 60, 20)) +
  geom_text(data = subset(Females_AllAges_0814, is.na(adj.RD.suicide)==F), 
            aes(x = adjusted.suicide.white + 1, label = ifelse(ST != "DC", paste0("+", round(-adj.RD.suicide)), round(-adj.RD.suicide))), size = 3, col = "#fd8d3c") +
  facet_wrap(~type)
```

##### PAGE BREAK

###Figure 3: The relationship between the rate of homicide and suicides among whites males by state, 2008-2014

```{r Figure3, fig.width=6, fig.height=6, echo = F, warning=F}
ggplot(subset(Males_AllAges_0814, type == "Firearm" & ST != "DC"), aes(adjusted.suicide.white, adjusted.homicide.white)) + 
  geom_point(aes(fill = gun.house.prev.category, size = average.pop.white), col = "black", pch = 21, alpha = 0.9) +
  geom_text_repel(aes(label = ST), alpha = 0.8, size = 3) +
  scale_fill_viridis(discrete = T, direction = -1) +
  ylab("White Homicide Rate\n(per 100,000)") +
  xlab("White Suicide Rate (per 100,000)") + 
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_size_area(breaks = c(100000, 500000, 1500000, 3000000, 7000000), max_size = 15, guide = F) +
  guides(fill = guide_legend(title = "Gun ownership", override.aes = list(size = 5)))

# ggplot(subset(Males_AllAges_0814, type == "Firearm"), aes(adjusted.suicide.white, adjusted.homicide.white)) + 
#   geom_point(aes(fill = gun.level, size = average.pop.white), col = "white", pch = 21, alpha = 0.5) +
#   geom_text_repel(aes(label = ST), alpha = 0.8, size = 3) +
#   ylab("White Homicide Rate\n(per 100,000)") +
#   xlab("White Suicide Rate (per 100,000)") + 
#   theme_bw() +
#   scale_size_area(breaks = c(100000, 500000, 1500000, 3000000, 7000000), max_size = 15, guide = F) +
#   guides(fill = guide_legend(title = "Gun ownership")) 
```

##### PAGE BREAK

###Figure 4: The relationship between the rate of homicide and suicides among black males by state, 2008-2014

```{r figure4, fig.width = 6, fig.height = 6, echo=F, warning=F}
ggplot(subset(Males_AllAges_0814, type == "Firearm" & ST != "DC"), aes(x = adjusted.suicide.black, y = adjusted.homicide.black)) + 
  geom_point(aes(fill = gun.house.prev.category, size = average.pop.black), col = "black", pch = 21, alpha = 0.9) +
  geom_text_repel(aes(label = ST), alpha = 0.8, size = 3, check_overlap = T) +
  scale_fill_viridis(discrete = T, direction = -1) +
  ylab("Black Homicide Rate\n(per 100,000)") +
  xlab("Black Suicide Rate (per 100,000)") + 
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_size_area(breaks = c(16600, 100000, 500000, 1400000), max_size = 15, guide = F) +
  guides(fill = guide_legend(title = "Gun ownership", override.aes = list(size = 5))) 
    
# ggplot(subset(Males_AllAges_0814, type == "Firearm"), aes(adjusted.suicide.black, adjusted.homicide.black)) + 
#   geom_point(aes(fill = gun.level, size = average.pop.black), col = "white", pch = 21, alpha = 0.5) +
#   geom_text_repel(aes(label = ST), alpha = 0.8, size = 3, check_overlap = T) +
#   ylab("Black Homicide Rate\n(per 100,000)") +
#   xlab("Black Suicide Rate (per 100,000)") + 
#   theme_bw() +
#   scale_size_area(breaks = c(16600, 100000, 500000, 1400000), max_size = 15, guide = F) +
#   guides(fill = guide_legend(title = "Gun ownership"))

# ggplot(subset(Males_AllAges_0814, type == "Firearm"), aes(x = gun.house.prev, y = adj.RD.homicide)) +
#          geom_point() +
#   geom_text_repel(aes(label = ST), alpha = 0.8, size = 3, check_overlap = T) 
# 
# ggplot(subset(Males_AllAges_0814, type == "Firearm"), aes(x = gun.house.prev, y = adj.RD.suicide)) +
#          geom_point() +
#   geom_text_repel(aes(label = ST), alpha = 0.8, size = 3, check_overlap = T) +
#   geom_smooth()
# 
# ggplot(subset(Males_AllAges_0814, type == "Firearm"), aes(x = gun.house.prev, y = Population.white)) +
#          geom_point() + geom_text_repel(aes(label = ST), alpha = 0.8, size = 3, check_overlap = T) 
# 
# ggplot(subset(Males_AllAges_0814, type == "Firearm"), aes(x = gun.house.prev, y = Population.black)) +
#          geom_point() + geom_text_repel(aes(label = ST), alpha = 0.8, size = 3, check_overlap = T) 
```

#####PAGE BREAK

###References