---
title: "Inequalities in violent death among men, by state, race, and firearms use"
output:
  word_document:
    reference_docx: reference.docx
bibliography: biblio.bib
csl: jama.csl
---

Corinne A Riddell, PhD

Sam Harper, PhD

Magdalena Cerdá, PhD

Jay S Kaufman, PhD

CAR, SH, and JSK: Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada.

MC: Violence Prevention Research Program, Department of Emergency Medicine, University of California, Davis, USA.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>. Phone: +1-438-863-0402

Word Count: 2700

##### PAGE BREAK

#Abstract

**Background:** It is unknown whether and how much the difference in excess deaths due to homicide among black men and suicide among white men vary by US state. Such knowledge could inform social policy aimed towards reducing social disparities in health.

**Objective:** Estimate the absolute inequality between non-Hispanic black and white men in firearm- and non-firearm-related homicide and suicide across US states.

**Design:** Surveillance study.

**Setting:** 50 states and the District of Columbia, 2008 to 2014.

**Participants:** Non-Hispanic black and non-Hispanic white males, all ages.

**Measurements:** Absolute risks and risk differences in firearm- and non-firearm-related homicide and suicide for black and white men.

**Results:** 71,217 homicides and 191,909 suicides were included in this analysis. Black-white differences in rates of firearm-related homicide and suicides varied widely across states. Relative to white men, black men experienced between 5 and 51 additional homicides per 100,000 men per year, with black men in Indiana, Michigan, Missouri, Pennsylvania, and Wisconsin all having more than 40 additional homicides per 100,000. White men experienced between -2 and 16 additional suicides per 100,000, with the largest inequalities measured in Southern and Western states, and smallest inequalities in densely populated Northeastern states.

**Limitations:** Non-firearm suicides may be misclassified as unintentional injury deaths. This analysis focused on deaths only, and does not examine rates and inequalities in non-fatal violence.

**Conclusion:** The large variation in the risks of violence and their racial inequalities highlights states where policies are most urgently needed to reduce violence and their racial inequalities. Future research linking these variations with the social, cultural, and policy environments are needed to identify settings that support reduced rates of violence for all Americans.

##### PAGE BREAK

```{r, echo=F, warning=F, message=F}
library(tidyverse)
library(ggrepel)
library(viridis)
```

```{r gun_ownership_data, echo=F, warning=F, message=F}
gun.house.prev <- c(52.2, 59.8, 32.3, 58.8, 20.1, 34.6, 18.1, 26.3, 25.2,
                    40.3, 10.2, 55.7, 20.7, 38.5, 45.7, 42.8, 47.7, 45.0,
                    40.3, 21.7, 11.5, 40.8, 41.2, 54.6, 44.2, 62.6, 45.4,
                    34.0, 31.0, 11.4, 39.7, 18.5, 39.4, 56.2, 34.0, 46.5,
                    39.8, 35.1, 12.4, 43.3, 59.9, 46.6, 37.1, 44.8, 43.8,
                    37.5, 34.0, 58.5, 43.0, 65.5)

gun.data <- data.frame("ST" = state.abb, "gun.house.prev" = gun.house.prev)

#ggplot(gun.data, aes(gun.house.prev)) + geom_histogram(bins = 20, col = "white") 
#ggplot(gun.data, aes(gun.house.prev)) + geom_density()
#quantile(gun.data$gun.house.prev, c(0.33, 0.66))

gun.data <- gun.data %>% arrange(gun.house.prev) %>% 
  mutate(gun.house.prev.category = cut(gun.house.prev, 
                                       breaks = c(0, 20, 35, 44.9, 66), 
                                       labels = c("10.2%-19.9%", "20.0%-34.9%", "35.0%-44.9%", "45.0%-65.5%")))
```

```{r load_data_age_std_males, echo=F, warning=F, message=F}
#homicide
NHWM_homicide_FA <- read.table("./Data/Men/Homicide_Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHWM_homicide_NonFA <- read.table("./Data/Men/Homicide_Non-Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_homicide_FA <- read.table("./Data/Men/Homicide_Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_homicide_NonFA <- read.table("./Data/Men/Homicide_Non-Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

#suicide
NHWM_suicide_FA <- read.table("./Data/Men/Suicide_Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHWM_suicide_NonFA <- read.table("./Data/Men/Suicide_Non-Firearms_White_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_suicide_FA <- read.table("./Data/Men/Suicide_Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")

NHBM_suicide_NonFA <- read.table("./Data/Men/Suicide_Non-Firearms_Black_NH_Males_Age-adjusted_2008-2014.txt", skip = 21, header = T, sep = ",")
```

```{r tidy_fx, echo=F, warning=F, message=F}
tidy_fx <- function(data, state, death, population, crude, adjusted, race, intent, mechanism) {
  dat2 <- data[, c(state, death, population, crude, adjusted)]
  names(dat2)[2] <- paste0(death, ".", intent, ".", race)
  names(dat2)[3] <- paste0(population, ".", race)
  names(dat2)[4] <- paste0("crude.", intent, ".", race)
  names(dat2)[5] <- paste0("adjusted.", intent, ".", race)
  dat2$type = mechanism
  
  return(dat2)
}
```

```{r tidy_male_data, echo=F}
NHWM_homicide_FA <- tidy_fx(data = NHWM_homicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "homicide", "Firearm")
NHWM_homicide_NonFA <- tidy_fx(data = NHWM_homicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "homicide", "Non-Firearm")
NHWM_suicide_FA <- tidy_fx(data = NHWM_suicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "suicide", "Firearm")
NHWM_suicide_NonFA <- tidy_fx(data = NHWM_suicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "white", "suicide", "Non-Firearm")

NHBM_homicide_FA <- tidy_fx(data = NHBM_homicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "homicide", "Firearm")
NHBM_homicide_NonFA <- tidy_fx(data = NHBM_homicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "homicide", "Non-Firearm")
NHBM_suicide_FA <- tidy_fx(data = NHBM_suicide_FA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "suicide", "Firearm")
NHBM_suicide_NonFA <- tidy_fx(data = NHBM_suicide_NonFA, "ST", "Deaths", "Population", "U_C_Rate", 
                            "U_A_Rate", "black", "suicide", "Non-Firearm")

NHBM_homicide<- rbind(NHBM_homicide_FA, NHBM_homicide_NonFA)
NHBM_suicide <- rbind(NHBM_suicide_FA, NHBM_suicide_NonFA)
NHWM_homicide <- rbind(NHWM_homicide_FA, NHWM_homicide_NonFA)
NHWM_suicide <- rbind(NHWM_suicide_FA, NHWM_suicide_NonFA)
rm(NHBM_homicide_FA, NHBM_homicide_NonFA, NHBM_suicide_FA, NHBM_suicide_NonFA,
   NHWM_homicide_FA, NHWM_homicide_NonFA, NHWM_suicide_FA, NHWM_suicide_NonFA)
```

```{r merge_data_males, echo=F}
#merge all the data
Males_AllAges_0814 <- merge(NHBM_homicide, NHBM_suicide %>% select(-Population.black), by = c("ST", "type")) %>%
  mutate(crude.RD.hom.su.black = crude.homicide.black - crude.suicide.black,
         adj.RD.hom.su.black = adjusted.homicide.black - adjusted.suicide.black)

Males_AllAges_0814 <- merge(Males_AllAges_0814, NHWM_homicide, by = c("ST", "type")) %>%
  mutate(crude.RD.homicide = crude.homicide.black - crude.homicide.white,
         adj.RD.homicide = adjusted.homicide.black - adjusted.homicide.white)

Males_AllAges_0814 <- merge(Males_AllAges_0814, NHWM_suicide %>% select(-Population.white), by = c("ST", "type")) %>%
  mutate(crude.RD.suicide = crude.suicide.black - crude.suicide.white,
         adj.RD.suicide = adjusted.suicide.black - adjusted.suicide.white)

Males_AllAges_0814 <- Males_AllAges_0814 %>% 
  mutate(crude.RD.hom.su.white = crude.homicide.white - crude.suicide.white, 
         adj.RD.hom.su.white = adjusted.homicide.white - adjusted.suicide.white)
```

```{r add_factor_sorting_males, echo = F}
#used for plotting -- to arrange the states on the y-axis according to the magnitude of the black-white inequality
levels.firearm.homicide <- Males_AllAges_0814 %>% filter(type == "Firearm") %>%
  transmute(ST.order.RD.homicide = reorder(ST, adj.RD.homicide))

# Males_AllAges_0814 %>% 
#   filter(is.na(adjusted.homicide.black) == T & type == "Firearm") %>% 
#   select(ST, adjusted.homicide.white)

Males_AllAges_0814$ST.order.RD.homicide <- factor(Males_AllAges_0814$ST, 
                                                  levels = c("HI", "SD", "NH", "ND", "VT", "ID", "ME", "WY", "MT",
                                                             levels(levels.firearm.homicide$ST.order.RD.homicide)[1:42]))

levels.firearm.suicide <- Males_AllAges_0814 %>% filter(type == "Firearm") %>%
  transmute(ST.order.RD.suicide = reorder(ST, -adj.RD.suicide))

# Males_AllAges_0814 %>% 
#   filter(is.na(adjusted.suicide.black) == T & type == "Firearm") %>% 
#   select(ST, adjusted.suicide.white)

Males_AllAges_0814$ST.order.RD.suicide <- factor(Males_AllAges_0814$ST, 
                                                  levels = c("RI", "HI", "IA", "NH", "ME", "SD", "ND", "VT", "UT", "ID", "MT", "WY",
                                                             levels(levels.firearm.suicide$ST.order.RD.suicide)[1:39]))

```

```{r add_variables_males, echo=F}
num.years = 2014-2008+1

Males_AllAges_0814 <- Males_AllAges_0814 %>% 
  mutate(average.pop.white = Population.white/num.years,
         average.pop.black = Population.black/num.years)

Males_AllAges_0814$gun.level <- "Middle"
Males_AllAges_0814$gun.level[Males_AllAges_0814$ST %in% c("LA", "UT", "OK", "IA", "TN","KY", "AL", 
                                                          "MS", "ID","ND", "WV", "AR", "AK", "SD", "MT", "WY")] <- "High"
Males_AllAges_0814$gun.level[Males_AllAges_0814$ST %in% c("HI", "NJ", "MA", "RI", "CT", "NY")] <- "Low"

Males_AllAges_0814$gun.level <- factor(Males_AllAges_0814$gun.level, levels = c("Low", "Middle", "High"))

#dim(Males_AllAges_0814)
Males_AllAges_0814 <- merge(Males_AllAges_0814, gun.data, by = "ST", all.x = T)
#dim(Males_AllAges_0814)
```

#Introduction

```{r WISQARS-data-intro, echo=F}
#source: WISQARS fatal injury mapping tool
#note:I used homicide, not homicide and legal intervention -- this changes the numbers slightly, not materially. 

#HOMICIDE relative risks
nh.black.homicide.rate.agestd <- 33.58 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.homicide.rate.crude <- 35.53 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.homicide.rate.agestd <- 3.46 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.homicide.rate.crude <- 3.45 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.homicide.black.white.agestd <- nh.black.homicide.rate.agestd/nh.white.homicide.rate.agestd
rr.homicide.black.white.crude <- nh.black.homicide.rate.crude/nh.white.homicide.rate.crude

#SUICIDE relative risks
nh.black.suicide.rate.agestd <- 9.64 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.suicide.rate.crude <- 9.32 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.suicide.rate.agestd <- 24.49 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.suicide.rate.crude <- 26.01 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.suicide.white.black.agestd <- nh.white.suicide.rate.agestd/nh.black.suicide.rate.agestd
rr.suicide.white.black.crude <- nh.white.suicide.rate.crude/nh.black.suicide.rate.crude

```

```{r WISQARS-data-intro-2, echo=F}

risk.homicide.nh.male <- 8.33 #per 100k, non-hispanic blacks + whites, 2008-2014, males
risk.homicide.nh.male.firearm <- 6.17 #per 100k, non-hispanic blacks + whites, 2008-2014, males
prop.homicide.nh.male.firearm <- (risk.homicide.nh.male.firearm/risk.homicide.nh.male)*100

risk.suicide.nh.male <- 22.42 #per 100k, non-hispanic blacks + whites, 2008-2014, males
risk.suicide.nh.male.firearm <- 12.87 #per 100k, non-hispanic blacks + whites, 2008-2014, males
prop.suicide.nh.male.firearm <- (risk.suicide.nh.male.firearm/risk.suicide.nh.male)*100
```

```{r WISQARS-data-intro-3, echo=F}
#source: WISQARS fatal injury mapping tool
#note:I uses homicide, not homicide and legal intervention -- this changes the numbers slightly, not materially. 

#FIREARM HOMICIDE relative risks
nh.black.homicide.rate.agestd.fa <- 27.58 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.homicide.rate.crude.fa <- 29.57 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.homicide.rate.agestd.fa <- 2.04 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.homicide.rate.crude.fa <- 2.01 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.homicide.black.white.agestd.fa <- nh.black.homicide.rate.agestd.fa/nh.white.homicide.rate.agestd.fa
rr.homicide.black.white.crude.fa <- nh.black.homicide.rate.crude.fa/nh.white.homicide.rate.crude.fa

#FIREARM SUICIDE relative risks
nh.black.suicide.rate.agestd.fa <- 5.33 #per 100k, annualized age standardized, 2008-2004, entire US, black, non-hispanic, males
nh.black.suicide.rate.crude.fa <- 5.04 #per 100k, annualized crude, 2008-2004, entire US, black, non-hispanic, males

nh.white.suicide.rate.agestd.fa <- 14.06 #per 100k, annualized age standardized, 2008-2004, entire US, white, non-hispanic, males
nh.white.suicide.rate.crude.fa <- 15.27 #per 100k, annualized crude, 2008-2004, entire US, white, non-hispanic, males

rr.suicide.white.black.agestd.fa <- nh.white.suicide.rate.agestd.fa/nh.black.suicide.rate.agestd.fa
rr.suicide.white.black.crude.fa <- nh.white.suicide.rate.crude.fa/nh.black.suicide.rate.crude.fa

```

Non-Hispanic black males are nearly `r round(rr.homicide.black.white.agestd)` times more likely to die from homicide than non-Hispanic white males in the United States, whereas white males are `r round(rr.suicide.white.black.agestd,1)` times more likely to die from suicide.[@wisqars_data] `r round(prop.homicide.nh.male.firearm)`% of these homicides and `r round(prop.suicide.nh.male.firearm)`% of these suicides involved firearms, and the relative racial inequalities in violent deaths are especially marked for firearm-related homicide and slightly increased for firearm-related suicide; Black males are `r round(rr.homicide.black.white.agestd.fa,1)` times more likely to die of a firearm-related homicide than white males, whereas white males are `r round(rr.suicide.white.black.agestd.fa,1)` times as likely to die from firearm-related suicide. [@wisqars_data] 

While racial inequalities in violent deaths are well-established for the United States overall, it is unknown whether and how much they vary by state. Large variation in racial inequalities among spatially contiguous or otherwise similar states may reflect differences in the social, cultural, and policy environments worthy of further investigation. Quantifying state-level heterogeneity is the first step towards implementing targeted policies aimed at reducing violence and inequalities where they are most urgently needed.

Across the US, states with the highest levels of household gun ownership have higher rates of firearm-related suicide and homicide.[@webster_book_table] Ecologic studies comparing US states have estimated positive associations between rates of gun ownership and firearm-related homicide and suicide rates,[@miller2002rates][@miller2007state][@siegel2013relationship] [@miller2013firearms][@miller2007household] and case-control studies have found positive associations between having a gun in the home and an individual’s risk of homicide occurring at home,[@kellermann1993gun] and with the overall risk of suicide.[@kellermann1992suicide] When considering the causal pathway to homicide or suicide, two component causes are motivation and opportunity.[@kubrin2009explaining] Gun availability or ownership increases opportunity and it is plausible that motivation varies by racial group. Thus, the relationship between gun availability and violence may be differential by race. This information is currently unknown but is important to inform whether policies aimed at reducing gun prevalence may have differential effects among blacks and whites.

In this study, we aimed to: (1) Estimate the absolute inequality between non-Hispanic black and white men in firearm- and non-firearm-related homicide and suicide across US states; (2) Examine the relationships between state gun ownership prevalence, firearm-related homicide, and firearm-related suicide, separately for black and white men.

#Methods

We used the Center for Disease Control and Prevention’s Web-based Injury Statistics Query and Reporting System (WISQARS) Fatal Injury Mapping tool to abstract the age-adjusted number of firearm fatalities for males in every state during 2008 to 2014 for non-Hispanic whites and non-Hispanic blacks, by intent (homicide or suicide), and firearms involvement. Throughout this paper we refer to this population as men since the vast majority of deaths occurred after childhood. The data were suppressed by WISQARS for race-state strata with 10 or fewer actual deaths between 2008 and 2014, and could not be used in these analyses. We computed the average annual fatality rate according to race, intent, and firearms involvement by dividing the number of fatalities by the total population estimate (i.e., the summation of population estimates over 2008 to 2014) and we calculated the absolute inequality (risk difference) in estimates between black and white men. For homicide, we subtract the fatality rate of white men from that of black men, and for suicide these rates are reversed, such that the estimated risk differences are mostly positive and interpreted as the number of additional deaths to black (white) men due to homicide (suicide) associated with the heightened fatality rate experienced by black (white) men. We did not calculate confidence intervals because there was no sampling, and thus no sampling error. To reflect varying population sizes among states and by race, we size the data markers according to race-specific average annualized population size. 

To examine the relationships between gun prevalence, firearm-related homicide, and firearm-related suicide by race, we used the last available state-level estimates of household firearm ownership from the National Behavioral Risk Factor Surveillance System (BRFSS) from 2004. The BRFSS is an annual telephone survey on the non-institutionalized adult population that asks respondents about their health-related risk behaviors, chronic health conditions, and use of preventive services. In 2001, 2002, and 2004, the BRFSS asked, “Are any firearms kept in or around your home?” 293,882 people responded to this question and estimates of household gun ownership prevalence (%) by state were computed previously and ranged between 10.2% in Hawaii and 65.5% in Wyoming.[@monuteaux2015firearm] Given that individuals may under-report gun ownership, and that state gun prevalence may have shifted since 2004, we grouped states into four broad categories of gun prevalence: (i) 10% - 19.9%, (ii) 20% - 34.9%, (iii) 35% - 44.9%, and (iv) 45% - 66%, to provide a fair proxy for current levels of ownership. 

We then visually assessed the relationships between state gun prevalence, firearm-related homicide, and firearm-related suicide by graphing the rate of firearm-related homicide as a function of firearm-related suicide and shaded the points by category of gun prevalence, separately for black men and white men. Correlation between rates of firearm-related homicide and suicide would imply that conditions that generate variation in homicide also generate variation in suicide, such as cultural, social, or economic conditions, including state gun prevalence or the firearm policy environment. Thus, the purpose of this descriptive exercise is to illustrate whether states with relatively high rates of firearm-related homicide also have relatively high rates of firearm-related suicide, and if this pattern exists for both black men and white men and is associated with state gun ownership. 

The analysis was performed in R version 3.3.2.[@r_language] No institutional ethical review was required since this study used publicly accessible data.[@tcps2] A replication data set including the raw data and statistical code to reproduce the manuscript is publicly available on GitHub: [https://github.com/corinne-riddell/InjuryDisparity/](https://github.com/corinne-riddell/InjuryDisparity/).

**Funding Source**

This research was not formally supported by a research grant, although one of the co-authors was supported by the National Institutes of Health (#1R21 DA041154-01).

#Results

```{r results_statistics, echo=F}
stats <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adj.RD.homicide, na.rm = T)), max = round(max(adj.RD.homicide, na.rm = T)))

stats2 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.homicide.white, na.rm = T)), max = round(max(adjusted.homicide.white, na.rm = T)))

white.pop.dc <- Males_AllAges_0814 %>% filter(ST == "DC" & type == "Firearm") %>% select(Population.white)

stats3 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.homicide.black, na.rm = T)), max = round(max(adjusted.homicide.black, na.rm = T)))

tib1 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(sum.deaths.homicide.black = sum(Deaths.homicide.black, na.rm = T), 
                                                    sum.deaths.homicide.white = sum(Deaths.homicide.white, na.rm = T), 
                                                    sum.deaths.suicide.black = sum(Deaths.suicide.black, na.rm = T), 
                                                    sum.deaths.suicide.white = sum(Deaths.suicide.white, na.rm = T))


```

There were `r formatC(sum(tib1$sum.deaths.homicide.black) + sum(tib1$sum.deaths.homicide.white), format = "d", big.mark = ",")` homicides and `r formatC(sum(tib1$sum.deaths.suicide.black) + sum(tib1$sum.deaths.suicide.white), format = "d", big.mark = ",")` suicides during 2008 to 2014 included in this analysis. Homicide deaths were suppressed for black men in 11 states (Alaska (non-firearm only), New Mexico (non-firearm only) Montana, Wyoming, Maine, Idaho, Vermont, North Dakota, New Hampshire, South Dakota, and Hawaii), and for white men in DC. Suicide deaths were suppressed in 12 states for black men (Montana, Wyoming, Maine, Idaho, Vermont, North Dakota, New Hampshire, South Dakota, Hawaii (firearm only), Utah (firearm only), Iowa (firearm only), and Rhode Island (firearm only)).

**Figure 1** displays the rates of homicide among black men compared to white men, according to state and firearms use. The black-white difference in homicide fatality rates varied strikingly by state, between `r stats$min[1]` and `r stats$max[1]` per 100,000 per year, and was driven by marked variation in firearm homicides to black men across states. States with the smallest black firearm homicide rates saw fewer than 15 deaths per 100,000 per year over the time period, compared to more than 40 deaths per 100,000 per year in states with the highest rates. Homicide rates for white men showed relatively little variation, between `r stats2$min[1]` and `r stats2$max[1]` deaths per 100,000 per year. The states with the highest rates of firearm homicide for black men (e.g., Missouri, Michigan, Illinois, Indiana) also had the largest differences in rates between blacks and whites, signaling settings that were most dangerous for blacks but relatively safe for whites. The District of Columbia also had a high rate of firearm homicide for blacks, but its rate for white men was suppressed. Given its abstracted cumulative population denominator of `r formatC(white.pop.dc$Population.white, format  = "d", big.mark = ",")`, this implies a rate $\leq$ 10 per `r formatC(white.pop.dc$Population.white, format  = "d", big.mark = ",")`, or `r round((10/white.pop.dc)*100000, 1)` per 100,000, implying that DC was also relatively dangerous for black men, but not for white men.

Georgia, New York, Florida, Texas, and California are the states with the largest black populations. Among them, Texas exhibited the smallest black-white difference in firearm-related homicide (13 additional deaths to black men per 100,000), which was also substantially lower than the differences estimated for neighboring states like Louisiana (+37), Oklahoma (+29), and Arkansas (+29). Georgia (+16) and Florida (+22) had differences close to nearby South and North Carolina (+19 and +17, respectively), while having lower rates than Alabama (+27) and Tennessee (+28). New York's difference (+18) was similar to Massachusetts (+17) and Connecticut (+22), but was substantially lower than the rates of neighboring New Jersey (+30) and Pennsylvania (+41). California exhibited the largest black-white difference (+30) in firearm-related homicide among the five states with the largest black populations, which was larger than the rates estimates for contiguous states like Nevada (+17) and Oregon (+10), although these states had relatively small black populations.

Although non-firearm homicides are much more rare, they also demonstrated the general pattern of heightened rates among black men compared to white men. Whereas black men experienced annual rates that varied between `r stats3$min[2]` and `r stats3$max[2]` per 100,000 by state, white rates varied between `r stats2$min[2]` and `r stats2$max[2]` per 100,000. Across the US, firearm homicide rates were higher than non-firearm rates for black men, and were more than 5.3 times higher in all states in the East North Central region (Wisconsin, Illinois, Indiana, Ohio, and Michigan), and in Missouri, Louisiana, Delaware, and Nebraska.

```{r results_statistics_suicide, echo=F}
stats4 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adj.RD.suicide, na.rm = T)),
                                                    max = round(max(adj.RD.suicide, na.rm = T)))

stats5 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.suicide.white, na.rm = T)),
                                                    max = round(max(adjusted.suicide.white, na.rm = T)))

stats6 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.suicide.black, na.rm = T)),
                                                    max = round(max(adjusted.suicide.black, na.rm = T)))

stats7 <- Males_AllAges_0814 %>% group_by(type) %>% summarise(min = round(min(adjusted.suicide.white, na.rm = T)),
                                                    max = round(max(adjusted.suicide.white, na.rm = T)))
```

In contrast to homicide, suicide fatality rates were higher among white men compared to black men, independent of firearms involvement, with the exception of firearm-related suicide in DC (**Figure 2**). The white-black difference in suicides involving a firearm varied from `r -(stats4$max[1])` per 100,000 in the District of Columbia to `r -(stats4$min[1])` per 100,000 per year in Missouri, although differences could not be estimated for 12 states with 10 or fewer firearm-related suicides among blacks. The annual rate of suicide involving firearms varied more markedly for white men than for black men: between `r stats5$min[1]` per 100,000 in DC and `r stats5$max[1]` per 100,000 in Wyoming for white men, compared to a range of `r stats6$min[1]` per 100,000 in Massachusetts to `r stats6$max[1]` per 100,000 in New Mexico among black men. 

Seven of the ten states with the largest white-black differences in firearm-related suicide are located in the South, including Mississippi, Arkansas, West Virginia, Alabama, Texas, South Carolina, and Louisiana, while the remaining three are in the West, including Nevada, Arizona, and Colorado (+12 and 16 per 100,000 per year). Along with DC, the mid-Atlantic states of Massachusetts, New Jersey, Connecticut, and New York had the smallest white-black differences (<4 per 100,000 per year) in firearm-related suicide. Importantly, these states also tended to have the lowest firearm-related suicide rates for both black men and white men, implying that the safest states (in terms of suicide risk) for black men and white men are the same. 

The rates of black- and white- non-firearm-related suicides and the difference between them varied substantially less than those involving firearms. For white men, between `r stats7$min[2]` per 100,000 in Alaska and `r stats7$max[2]` per 100,000 in Hawaii non-firearm-related suicides were committed each year on average across the states, compared to `r stats6$min[2]` per 100,000 in Alabama and `r stats6$max[2]` per 100,000 in Utah among black men.

The five states with jointly low rates of firearm-related homicide and suicide among white men (i.e., Massachusetts, New Jersey, Rhode Island, New York, and Connecticut) were also in the lowest category of reported gun prevalence (**Figure 3**). Furthermore, many of the states with the highest violence rates belonged to the highest category of gun prevalence, while most states located in the middle of the graph belonged to the middle categories of gun prevalence. Notably, several states that belonged to the highest gun prevalence category had relatively low rates of firearm-related homicide, including Iowa, Nebraska, the Dakotas, Idaho, Montana, and Wyoming. Generally speaking, the strong linear relationship between firearm-related homicide and suicide among white men suggests that the characteristics that generate variation in homicide and suicide in this racial/ethnic group were similar.

Among black men, states with the lowest reported prevalence of guns had the lowest rates of firearm-related suicide, and below-average rates of firearm-related homicide (**Figure 4**). States with the highest rates of violence (i.e., Mississippi, Missouri, Oklahoma) had heightened rates of gun prevalence, although several Southern states with relatively high gun prevalence had relatively low rates of firearm-related homicide. The weak relationship between firearm-related homicide and suicide among black men implies that the set of characteristics that generated variation in firearm-related homicide among black men are different from the characteristics that generated variation in firearm-related suicide.

#Discussion

Using data from all 50 US states and DC, our study illustrates striking variation in the inequality in firearm-related homicide. Black men experienced between 5 and 51 additional homicides per 100,000, and the magnitude of the difference was largely driven by differences among states in homicide rates for black men. Rust-belt states reported the most firearm homicides among black men and medium levels of gun ownership. In the South, states with high levels of gun ownership, such as Louisiana, reported higher homicide rates among black men than states with medium levels, such as Texas. Disentangling the factors that make several Southern states relatively safe for black men and the rust-belt states relatively unsafe compared to white men should be a high priority in order to reduce health inequalities and foster communities that offer equal safety independently of race. Sociologists have found that structural disadvantage (i.e., concentrated poverty, joblessness, family disruption),[@Wilson_book] racial residential segregation,[@Massey_book][@ousey2009exploring][@shihadeh1996segregation][@peterson1993racial] and participation in drug markets[@blumstein1995youth][@fryer2013measuring] are implicated in the higher risk of homicide among black men relative to white men. Thus, multilevel research could incorporate state factors (i.e., gun ownership, and gun policy) alongside these neighborhood and individual characteristics to describe how layers of the social environment affect homicide and inequality. 

White men were at increased risk for firearm-related suicide, and experienced between -2 and 16 additional suicides per 100,000 compared to black men. The range in the suicide rate across states for white men was large, although less than the range in the homicide rate for black men. State gun prevalence was strongly tied to firearm-related suicide for both black and white men, and is the most lethal means of attempted suicide.[@spicer2000suicide][@miller2004epidemiology] States that were the safest also have some of the country's most restrictive gun laws,[@siegel2017firearm] highlighting the bidirectional relationship between gun policy and gun ownership, and the constellation of factors associated with the lowest risks of suicide in the country. Alcohol is another important risk factor for suicide, both in terms of history of alcoholism and overuse at the time of suicide.[@moscicki1997identification][@moscicki2001epidemiology] Alcohol was found in the systems of 25% of white male suicides tested and 15% of black males suicides in 2005--2006, with 16% of the white men being described as alcohol dependent at the time of suicide compared to 7% of black men.[@alcohol_suicide] Given that white men have higher rates of alcohol abuse than black men,[@alcohol_suicide] alcohol use might be implicated both in the direction of the white-black inequality and its variation across states.  

This study is subject to several limitations. In terms of the listed cause of death used to identify suicides and homicides, non-firearm suicides may be mistakenly coded as unintentional injuries or vice versa, and it may be especially difficult to determine intent from drug or alcohol overdoses. We considered only fatal outcomes, not attempted suicides and homicides. Given that firearm suicide attempts are more than 80% fatal,[@miller2004epidemiology][@spicer2000suicide] and firearm homicide attempts are approximately 20% fatal,[@cook2017constant] studying only fatal outcomes misses the morbidity endured during nonfatal assaults and suicide attempts. We were limited to contrasting age-standardized mortality rates, but would have rather contrasted age-specific rates, had there been more deaths. We did not examine women because their risks of homicide and suicide are much lower than for men, and data for black women were suppressed for more than half of the states. Finally, these results are purely descriptive and make no claims regarding causal mechanisms. Our purpose was to describe how much inequalities in violence vary across the United States in order to provide a foundation for future research linking these differences to variations in the social, cultural, and environmental settings where black and white communities are based, including variations in firearm policy and access to firearms. 

Firearms feature differently in the deaths of black and white men in the United States. The large variation in the homicide and suicide rates and their racial differences across states places a spotlight on states where policies are most urgently needed to reduce rates of violence, and their racial inequalities. This research provides the first step in measuring these inequalities across states, which can be further studied to identify the policies, and social and cultural factors that support reduced rates of violence for all Americans.

##Figure Legends 
###Figure 1: Difference in age-standardized homicide fatalities rates for men by state, firearms, and race, 2008-2014

```{r figure1, fig.width=6, fig.height=8.5, echo = F, warning=F, eval=F}
ggplot(Males_AllAges_0814, aes(y = ST.order.RD.homicide, x = adjusted.homicide.white)) + 
  geom_segment(aes(x = adjusted.homicide.white, xend = adjusted.homicide.black, 
                   y = ST.order.RD.homicide, yend = ST.order.RD.homicide), 
               lwd = 2, col = "#8A5C7B", alpha = 0.5) +
  geom_point(aes(col = "NH White", size = average.pop.white)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1)
  geom_point(aes(x = adjusted.homicide.black, col = "NH Black", size = average.pop.black)) + #alpha = ifelse(death.homicide.black < 20, 0.7, 1),
  scale_color_manual(values = c("#8A5C7B", "#4FD5AF")) +
  scale_alpha_continuous(guide = F) +
  scale_size_area(breaks = c(10000, 100000, 500000, 1500000, 3000000, 7000000)) +
  ylab("") +
  xlab("Homicide mortality per 100k") + 
  theme_bw() + 
  guides(color = guide_legend(title = " "), size = guide_legend(title = "Population")) + 
  theme(legend.position = c(0.8, 0.25), 
        plot.title = element_text(size = 10),
        legend.background = element_rect(fill=alpha('white', 0.2)),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12)) + 
   geom_text(data = Males_AllAges_0814 %>% filter(is.na(adj.RD.homicide) == F),
             aes(x = adjusted.homicide.black + 8, label = paste0("+", round(adj.RD.homicide))),
             size = 3, col = "#754255") +
  facet_wrap( ~ type) +
  scale_x_continuous(limits = c(0, 70)) 
```

###Figure 2: Difference in age-standardized suicide fatalities rates for men by state, firearms, and race, 2008-2014

```{r figure2, fig.width=6, fig.height=8.5, echo = F, warning=F, eval=F}
ggplot(Males_AllAges_0814, aes(y = ST.order.RD.suicide, x = adjusted.suicide.white)) + 
  geom_segment(aes(x = adjusted.suicide.white, xend = adjusted.suicide.black, 
                   y = ST.order.RD.suicide, yend = ST.order.RD.suicide), 
               lwd = 2, col = "#4FD5AF", alpha = 0.5) +
  geom_point(aes(col = "NH White", size = average.pop.white)) + 
  geom_point(aes(x = adjusted.suicide.black, col = "NH Black", size = average.pop.black)) + 
  scale_color_manual(values = c("#8A5C7B", "#4FD5AF")) +
  scale_alpha_continuous(guide = F) +
  scale_size_area(breaks = c(10000, 100000, 500000, 1500000, 3000000, 7000000)) +
  ylab("") + xlab("Suicide mortality per 100k") + theme_bw() + 
  guides(color = guide_legend(title = " "), size = guide_legend(title = "Population")) + 
  theme(legend.position = c(0.90, 0.27), 
        plot.title = element_text(size = 10),
        legend.background = element_rect(fill=alpha('white', 0.2)),
        strip.background = element_rect(fill="white"), 
        strip.text.x = element_text(size = 12),
        strip.text.y = element_text(size = 12)) + 
  scale_x_continuous(limits = c(0, 35)) +
  geom_text(data = subset(Males_AllAges_0814, is.na(adj.RD.suicide)==F), 
            aes(x = adjusted.suicide.white + 4, label = ifelse(ST != "DC", paste0("+", round(-adj.RD.suicide)), round(-adj.RD.suicide))), size = 3, col = "#308E77") +
  facet_wrap(~type)
```

###Figure 3: The relationship between the annual rates of homicide and suicide among white men by state and reported firearm ownership, 2008-2014

```{r Figure3, fig.width=6, fig.height=6, echo = F, warning=F, eval = F}
ggplot(subset(Males_AllAges_0814, type == "Firearm" & ST != "DC"), aes(adjusted.suicide.white, adjusted.homicide.white)) + 
  geom_point(aes(fill = gun.house.prev.category, size = average.pop.white), col = "black", pch = 21, alpha = 0.9) +
  geom_text_repel(aes(label = ST), size = 4) +
  scale_fill_manual(values = c("#D2F172","#53D6AD", "#5FA6C5", "#996E94")) +
  ylab("White Homicide Rate\n(per 100,000)") +
  xlab("White Suicide Rate (per 100,000)") + 
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_size_area(breaks = c(100000, 500000, 1500000, 3000000, 7000000), max_size = 15, guide = F) +
  guides(fill = guide_legend(title = "Gun ownership", override.aes = list(size = 5)))
```

###Figure 4: The relationship between the annual rates of homicide and suicide among black men by state and reported firearm ownership, 2008-2014

```{r figure4, fig.width = 6, fig.height = 6, echo=F, warning=F, eval = F}
ggplot(subset(Males_AllAges_0814, type == "Firearm" & ST != "DC"), aes(x = adjusted.suicide.black, y = adjusted.homicide.black)) + 
  geom_point(aes(fill = gun.house.prev.category, size = average.pop.black), col = "black", pch = 21, alpha = 0.9) +
  geom_text_repel(aes(label = ST), size = 4, check_overlap = T) +
  scale_fill_manual(values = c("#D2F172","#53D6AD", "#5FA6C5", "#996E94")) +
  ylab("Black Homicide Rate\n(per 100,000)") +
  xlab("Black Suicide Rate (per 100,000)") + 
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_size_area(breaks = c(16600, 100000, 500000, 1400000), max_size = 15, guide = F) +
  guides(fill = guide_legend(title = "Gun ownership", override.aes = list(size = 5))) 
```

**Acknowledgements:** Corinne Riddell is a post-doctoral researcher at McGill University, and received salary support from the Department of Epidemiology, Biostatistics, and Occupational Health. Sam Harper is supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec. Magdalena Cerdá is a CAMPOS Faculty Scholar, ADVANCE Program at the University of California, Davis and supported by the National Institutes of Health (#1R21 DA041154-01).

###References

1. Statistics, Programming & Economics Branch, National Center for Injury Prevention & Control, Centers for Disease Control and Prevention. Fatal injury mapping. Available from: https://wisqars.cdc.gov:8443/cdcMapFramework/. [cited 3 Oct 2017].

2. Miller M, Azreal D, Hemenway D. Firearms and violent death in the United States. In: Webster DW, Vernick JS, eds. *Reducing Gun Violence in America. Informing Policy with Evidence and Analysis*. Baltimore: The Johns Hopkins University Press; 2013:14-15.

3. Miller M, Azrael D, Hemenway D. Rates of household firearm ownership and homicide across us regions and states, 1988–1997. *American Journal of Public Health*. 2002;92(12):1988-1993.

4. Miller M, Hemenway D, Azrael D. State-level homicide victimization rates in the us in relation to survey measures of household firearm ownership, 2001–2003. *Social Science & Medicine*. 2007;64(3):656-664.

5. Siegel M, Ross CS, King III C. The relationship between gun ownership and firearm homicide rates in the united states, 1981–2010. *American Journal of Public Health*. 2013.

6. Miller M, Barber C, White RA, Azrael D. Firearms and suicide in the United States: Is risk independent of underlying suicidal behavior? *American Journal of Epidemiology*. 2013;178(6):946-955.

7. Miller M, Lippmann SJ, Azrael D, Hemenway D. Household firearm ownership and rates of suicide across the 50 United States. *Journal of Trauma and Acute Care Surgery*. 2007;62(4):1029-1035.

8. Kellermann AL, Rivara FP, Rushforth NB, et al. Gun ownership as a risk factor for homicide in the home. *New England Journal of Medicine*. 1993;329(15):1084-1091.

9. Kellermann AL, Rivara FP, Somes G, et al. Suicide in the home in relation to gun ownership. *New England Journal of Medicine*. 1992;327(7):467-472.

10. Kubrin CE, Wadsworth T. Explaining suicide among blacks and whites: How socioeconomic factors and gun availability affect race-specific suicide rates. *Social Science Quarterly*. 2009;90(5):1203-1227.

11. Monuteaux MC, Lee LK, Hemenway D, Mannix R, Fleegler EW. Firearm ownership and violent crime in the us: An ecologic study. *American journal of Preventive Medicine*. 2015;49(2):207-214.

12. R Core Team. *R: A Language and Environment for Statistical Computing*. Vienna, Austria: R Foundation for Statistical Computing; 2014. http://www.R-project.org/.

13. Canadian Institutes of Health Research, National Sciences and Engineering Research Council of Canada, and Sciences Sciences and Humanities Research Council of Canada. *Tri-council policy statement: Ethical conduct for research involving humans*, December 2014.

14. Wilson WJ. *The Truly Disadvantaged: The Inner City, the Underclass, and Public Policy*. University of Chicago Press; 1987.

15. Massey D. *Problem of the Century*. New York: Russell Sage; 2001:314-344.

16. Ousey GC, Kubrin CE. Exploring the connection between immigration and violent crime rates in US cities, 1980–2000. *Social Problems*. 2009;56(3):447-473.

17. Shihadeh ES, Flynn N. Segregation and crime: The effect of black social isolation on the rates of black urban violence. *Social Forces*. 1996;74(4):1325-1352.

18. Peterson RD, Krivo LJ. Racial segregation and black urban homicide. *Social Forces*. 1993;71(4):1001-1026.

19. Blumstein A. Youth violence, guns, and the illicit-drug industry. *The Journal of Criminal Law and Criminology*. 1995;86(1):10-36.

20. Fryer RG, Heaton PS, Levitt SD, Murphy KM. Measuring crack cocaine and its impact. *Economic Inquiry*. 2013;51(3):1651-1681.

21. Spicer RS, Miller TR. Suicide acts in 8 states: Incidence and case fatality rates by demographics and method. *American Journal of Public Health*. 2000;90(12):1885.

22. Miller M, Azrael D, Hemenway D. The epidemiology of case fatality rates for suicide in the Northeast. *Annals of Emergency Medicine*. 2004;43(6):723-730.

23. Siegel M, Pahn M, Xuan Z, et al. Firearm-related laws in all 50 US states, 1991–2016. *American Journal of Public Health*. 2017;(0):e1-e8.

24. Mościcki EK. Identification of suicide risk factors using epidemiologic studies. *Psychiatric Clinics of North America*. 1997;20(3):499-517.

25. Mościcki EK. Epidemiology of completed and attempted suicide: Toward a framework for prevention. *Clinical Neuroscience Research*. 2001;1(5):310-323.

26. Disease Control C for, Prevention. Alcohol and suicide among racial/ethnic populations — 17 states, 2005–2006. *MMWR*. 2009;58:638-640.

27. Cook PJ, Rivera-Aguirre AE, Cerdá M, Wintemute G. Constant lethality of gunshot injuries from firearm assault: United states, 2003–2012. *American Journal of Public Health*. 2017;107(8):1324-1328.

```{r echo=F, eval=F}
save(Males_AllAges_0814, file = "./InjuryDisparity/Data/Males_AllAges_0814.Rdata")
```

